Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-07-07 analysis/paper/paper.qmd
stash

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.

@@

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code. In most cases this can be attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change the way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.

>>>>

The PCA results in @fig-pca show that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.

@@

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code. In most cases this can be attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change the way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.
